---
title: "Lesson 6 - ggplot2 Examples"
author: "DS705"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE}
require(DS705data)
require(ggplot2)
require(ggfortify) # adds some functionality for regression diagnostics
require(gridExtra)
```

Setup the regression data from the presentation:
```{r}
production = data.frame(
  NumItems = c(175,189,344, 88,114,338,271,173,284,277,
               337, 58,146,277,123,227, 63,337,146, 68),
  RunTime =  c(195,215,243,162,185,231,234,166,253,196,
               220,168,207,225,169,215,147,230,208,172)
)
```

# Basic Scatterplots with regression lines

The code for these plots was adapted from http://www.cookbook-r.com/Graphs/Scatterplots_(ggplot2)/

```{r fig.height = 3, fig.width = 4}

ggplot( data = production, aes( x = NumItems, y = RunTime) ) + 
  geom_point( shape = 1) # use hollow circles

ggplot( data = production, aes( x = NumItems, y = RunTime) ) + 
  geom_point( shape =  1) + 
  geom_smooth( method = lm ) # add linear regression line includes 95% confidence band by default

ggplot( data = production, aes( x = NumItems, y = RunTime) ) + 
  geom_point( shape =  1) + 
  geom_smooth( method = lm, # add linear regression line
               se = FALSE ) # remove confidence bands

ggplot( data = production, aes( x = NumItems, y = RunTime) ) + 
  geom_point( shape =  1) + 
  geom_smooth() # add loess smoothed fit with confidence region
# loess stands for local regression and uses nearest-neigbors 
# to achieve a local fit

ggplot( data = production, aes( x = NumItems, y = RunTime) ) + 
  geom_point( shape =  1) + 
  geom_smooth( span = .5 ) # change the smoothness parameter for less smoothing
```                             

We don't cover local regression or `loess` in this class, but you're likely to see it around.  I really like this simple explanation of LOESS smoothing and the role of the span parameter:  https://stackoverflow.com/a/44242805.

## Adding a prediction interval band

`ggplot2` doesn't have this built in so you'll have to add it yourself:

```{r, message = FALSE, fig.height=3, fig.width=4}
# sort the dataframe by NumItems so that predictions will be 
# in the correct order for plotting
production <- production[ order( production$NumItems ),]

model <- lm( RunTime ~ NumItems, data = production)
# bind the predicted values to the dataframe
productionNew <- cbind( production, predict( model, interval="prediction") )

ggplot( data = productionNew, aes( x = NumItems, y = RunTime ) ) + 
  geom_point() +
  geom_smooth(method=lm, se = TRUE) + 
  geom_ribbon( aes( ymin = lwr, ymax = upr), 
               fill = 'blue', alpha = 0.2)
```

## Linear Regression Diagnostics

Many of the plots that were discussed in the presentations are available by using `autoplot()` from the `ggfortify` package.

```{r}
# change width to a subset of 1 through 6, e.g. c(1,3,5) to see
# only some of the plots
autoplot( model, which = 1:6, ncol = 2, label.size = 3)
```

## Mutliple Regression Lines

There is probably a nicer way to put this together, but this shows how to reproduce the plots near the end of the second presentation.

```{r}
data(HealthExam)
model <- lm(Cholesterol~AgeGroup + SysBP + AgeGroup:SysBP,data=HealthExam)

ggplot( data = HealthExam, aes( x = SysBP, y = Cholesterol) ) +
  geom_point(aes( color = AgeGroup) ) + 
  scale_color_manual( values = c("red", "darkgreen", "blue") ) + 
  geom_abline( intercept = 524, slope = -2.39, color = "red" ) + 
  geom_abline( intercept = -875.6, slope = 10.65, color = "darkgreen" ) + 
  geom_abline( intercept = 116.7, slope = 1.82, color = "blue" )
```

Here are the plots separated by age group.

```{r fig.height = 3, fig.width=9}
plot1 <- ggplot( data = HealthExam[HealthExam$AgeGroup=="18 to 35",], 
        aes( x = SysBP, y = Cholesterol) ) +
  geom_point( color = "red" ) + 
  geom_abline( intercept = 524, slope = -2.39, color = "red" ) + 
  labs( title = "Age Group 18 to 35")

plot2 <- ggplot( data = HealthExam[HealthExam$AgeGroup=="36 to 64",], 
        aes( x = SysBP, y = Cholesterol) ) +
  geom_point( color = "darkgreen" ) + 
  geom_abline( intercept = -875.6, slope = 10.65, color = "darkgreen" ) +
  labs( title = "Age Group 36 to 64")

plot3 <- ggplot( data = HealthExam[HealthExam$AgeGroup=="65+",], 
        aes( x = SysBP, y = Cholesterol) ) +
  geom_point( color = "blue" ) + 
  geom_abline( intercept = 116.7, slope = 1.82, color = "blue" ) + 
  labs( title = "Age Group 65+")
  
grid.arrange(plot1,plot2,plot3,ncol=3)
```

There may be a more elegant way to produce the last few plots showing regression lines by group.  Let us know if you find it.