---
title: "Lesson 5 - ggplot2 Examples"
author: "DS705"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE}
require(DS705data)
require(ggplot2)
```

## Bar Graphs/Charts for a Single Categorical Variable

To make a bar chart where we specify the value or height of each bar we might do something like this:

```{r}
dat <- data.frame(
  loanType = factor(c("Mort","Auto","Credit","Other"),
                    levels=c("Mort","Auto","Credit","Other")),
  observed = c(24,21,6,4)
)
ggplot(data = dat, aes( x = loanType, y = observed, fill = observed) ) + 
  geom_bar( stat = "identity" ) + 
  labs( x = "Type of Loan", 
        y = "Number Observed", 
        title = "Loans Summary" )
```

If we want R to tabulate the counts for us and make a bar graph it might go like this:

```{r}
data("HealthExam") # from DS705data package
ggplot( data = HealthExam, aes( x = Region )) +
  geom_bar(stat="count")
ggplot( data = HealthExam, aes( x = Region, fill = Region )) +
  geom_bar(stat="count")
```

## Bar Graphs/Charts for two categorical variables

Here are some examples for plotting the distribution of Age Group within each region:

```{r}
ggplot( data = HealthExam, aes( x = Region, fill = AgeGroup) ) + 
  geom_bar(stat="count")

ggplot( data = HealthExam, aes( x = Region, fill = AgeGroup) ) + 
  geom_bar(stat="count",position=position_dodge())

ggplot( data = HealthExam, aes( x = Region, fill = AgeGroup) ) + 
  geom_bar(stat="count",position=position_dodge(), color="black")
```

The last one is our favorite for this example.  However, the default colors in ggplot2 aren't color-blind friendly as they have equal luminance.  See this article to learn about changing color palettes, in particular check out the information about the `RColorBrewer` package:
http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/

We'll close with an example of manually defining a color-blind friendly palette and using it for our last graph.  This palette comes from the website that we just mentioned.

```{r}
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

ggplot( data = HealthExam, aes( x = Region, fill = AgeGroup) ) + 
  geom_bar(stat="count",position=position_dodge(), color="black") +
    scale_fill_manual(values=cbPalette)
```