---
title: "Lesson 7 - ggplot2 Examples"
author: "DS705"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE}
require(DS705data)
require(ggplot2)
require(GGally)
```

```{r, message=FALSE, warning=FALSE}
data(farmpond)
rich.out <- glm(RICH~FISH + TOTNITR,data=farmpond,family="binomial")
xv <- seq(0,15,.1) # create a sequence of values for plotting
n <- length(xv)
newdata <- data.frame( FISH = c( rep(0,n), rep(1,n) ), 
                       TOTNITR = c( xv, xv) )
prob <- predict(rich.out, newdata, type = "response")
dfPlot = cbind( newdata, prob )
dfPlot$FISH = factor(dfPlot$FISH) # make FISH a factor for plotting distinct groups

ggplot() + 
  geom_line( data = dfPlot, aes( x= TOTNITR, y = prob, group = FISH, color = FISH ),
             size = 1, linetype = 'dashed') + 
  geom_point( data = farmpond, aes( x = TOTNITR, y = RICH), shape = 1 ) + 
  labs( x = 'Total Nitrogen (mg/L)', y = 'Prob. of Species Richness at least 4')
```

```{r}
newdata0 <- data.frame( FISH = rep(0,n), TOTNITR = xv )
raw <- predict(rich.out, newdata0, se.fit=TRUE) 
prob0 <- exp(raw$fit)/(1+exp(raw$fit))
# define the level of confidence
C = .95
# get the appropriate critical value for the given level of confidence
crit = qnorm(1-(1-C)/2)
# lower confidence band
lwr = exp(raw$fit-crit*raw$se.fit)/(1+exp(raw$fit-crit*raw$se.fit))
# upper confidence band
upr = exp(raw$fit+crit*raw$se.fit)/(1+exp(raw$fit+crit*raw$se.fit))
# put it all in a dataframe for ggplot
dfPlot <- data.frame( xv, prob0, lwr, upr )
ggplot() + 
  geom_line( data = dfPlot, aes( x = xv, y = prob0), color = 'blue') + 
  geom_ribbon( data = dfPlot, aes( x = xv, ymin = lwr, ymax = upr), 
               fill = 'blue', alpha = 0.2) +
  geom_point( data = farmpond, aes( x = TOTNITR, y = RICH), shape = 1 ) +
  labs( x = 'Total Nitrogen (mg/L)', 
        y = 'Prob. of Species Richness at least 4',
        title = "Probabilities with 95% Confidence Bands (FISH = 0)")
```

In swirl we saw how to use `pairs()` to make a matrix of scatterplots for examinining the relationships between quantitative variables.  `ggplot2` doesn't have that built-in but the package `GGally` has some nice tools including a command called `ggpairs()`.  We've loaded `GGally` at the top of this file, but you may have to install it first.

```{r}
data(HealthExam)
ggpairs(HealthExam, mapping = aes( alpha = .4), columns = c("Weight", "SysBP","BodyMass","Cholesterol") )
```

We get some added stuff that `pairs()` doesn't provide including kernel density estimates for each quantitative variable and correlation coefficients.  You can also get more information if you have a factor variable like gender.  You can color using a factor variable like Sex in this case.

```{r}
data(HealthExam)
ggpairs(HealthExam, aes( alpha = .4, color = Sex), columns = c("Weight", "SysBP","BodyMass","Cholesterol") )
```

You could also add a factor variable as an additional column.  We've adde message=FALSE to the next R chunk to suppress some warnings about using 30 bins.

```{r message = FALSE}
data(HealthExam)
ggpairs(HealthExam, aes( alpha = .4, color = Sex), columns = c("Weight", "SysBP","BodyMass","Cholesterol","Sex") )
```

